A Survey of Possible Exoskeleton Control Architectures and Algorithms\\ (Draft 2.2)

### 2015-09-22 control.tex
cga updates

@@

\newcommand{\gc}{{\mbox {GravityCompensation}}}
\newcommand{\invdyn}{{\mbox {\tiny ID}}}

@@

This paper surveys possible exoskeleton control architectures and
algorithms.
One goal is to address the question:
{\it How do exoskeleton control approaches compare with 
approaches to humanoid robot control?}
A companion paper surveys implemented exoskeleton control architectures
and algorithms~\cite{}.

@@

A later white paper will discuss how to handle malfunctions and damage to the
system. Basically this boils down to fault detection, and switchover to
an approapriate safe mode of operation.

@@

\section{What are some design philosophy alternatives?}

\subsection{``Invisible'' exoskeleton}

Can we build an exoskeleton allows the operator to behave naturally
and exerts little or no force on the operator?

This could be achieved through active control using feedback of the
operator-exoskeleton forces. Additional gravity, friction, actuator, and
inertial compensation (inverse dynamics) can improve performance.

\subsection{``Natural'' exoskeleton}

Can we build an exoskeleton allows the operator to behave naturally?
The operator can feel the exoskeleton, but can overpower it when necessary.

Bypass valves and clutches could allow the operator to physically ``take over''
and push the relatively lightweight leg mounted exoskeleton around.

``Passive'' degrees of freedom can have physical or ``virtual'' springs
applied to assist in support (all abduct/adduct DOF, rotational DOF).
These aids can be clutched out or turned off when necessary.
``Virtual'' springs (springs implemented actively through exoskeleton
actuators) can be modulated.

\subsection{``Symbiotic'' exoskeleton}

Can we build an exoskeleton that allows the operator/exoskeleton team
to be effective, but not necessarily behave naturally? 
The operator needs to learn to ``fly'' the suit,
just as operators learn to operate parachutes,
wing suits, diving equipment, high and low temperature protective
suits, firefighting equipment, and high altitude flight
and parachute suits.

The exoskeleton can adapt its behavior to the task.
For example, for jumping, the exoskeleton can operate like a pogo 
stick:\\
\url{https://www.youtube.com/watch?v=lbp41vWP4o4},\\ 
and for running it could act like jumping stilts:\\
\url{https://www.youtube.com/watch?v=9ZOd7yEyhwI},

@@

{\bf Low Level Control:} Present interface to actuators and
high bandwidth feedback control, degrees of freedom,
or synergies of either:
\begin{itemize}
\item
Position control.
\item
Pure force or torque control.
\item
Impedance control: specify desired position (for spring), and generate
stiffness, damping, and potentially a modified inertia or moment of inertia.
Modifying inertia is more dangerous than providing active stiffness
or damping.
Or (almost equivalently),
specify a desired force and change the force according to position, velocity,
and acceleration.
\item
Online optimization of accelerations, actuator forces and torques, and contact forces
using quadratic programming.
\end{itemize} 
One important synergy is coordinated control of the center of mass location and
velocity for balance.
PID or fancier feedback control laws are implemented at this level.
Gravity, friction, actuator dynamics, and other feedforward disturbance
compensation are implemented at this level, as well as any inverse dynamics
or other types of feedforward reference tracking control.
Disturbance observers for model errors and contact/wind forces
are implemented at this level.
Typically this is done by a high sampling rate global servo~\cite{Atlas-robot}
or distributed limb or joint-level high sampling rate servos~\cite{Sarcos-robot}.
Often, this controller is provided by the manufacturer of the robot
or exoskeleton and runs on a separate computer system from the
``user's'' or task/application controller~\cite{Sarcos,Atlas}.

{\bf ``Phoneme'' Control:} Mid-level controllers
for relatively constant phases of behavior.
Often involves position or velocity targets (goto or
stay-at X), trajectory references, ...
Example from walking: left-stance, double-support-1, right-stance, double-support-2, ...;
Example from running: left-foot-down, flight-1, right-foot-down, flight-2, ...;
Example from horizontal jumping: hjump-pushoff, hjump-flight, hjump-prepare-landing, hjump-impact, hjump-balance, ...;
Finer resolution may be needed. For example a walking stance period may
be broken up into heel-strike, first-half, second-half, pushoff, and toe-off.
The first and second halves of stance are divided by when the center of mass
passes over the ankle.
Local inverse kinematics (finding small joint motions for small target changes)
is performed at this level.
This level often provides the behavioral primitives or ``verbs'' for robot programming.

{\bf ``Word'' Control:} Mid-level controllers for sequences of ``phoneme'' behavioral
units.
Often involves finite state machines, timers, if-then condition testing, gain
scheduling, ...
Examples: walk-to-location(x=7.8m, y=2.3m), run-to-location(speed=3.7m/s),
horizontal-jump(height= 2m, length= 5m).
Global inverse kinematics (selection of solution branch) is performed at this level.
This level often determines the control flow of a behavior.

{\bf Highest Level Control} aka {\bf Behavior Selection:} 
Answer the question: ``What do I do next?''
Select behaviors and behavioral parameters (targets,
speeds, durations).
Often involves finite state machines. timers, if-then condition testing, ...
This level often involves a great deal of human operator online interaction,
or is completely done by the operator.

{\bf Error Handling:} Errors must be detected and handled by parameter (target)
adjustment and/or behavior switching
at all of these levels.

{\bf State Estimation:}
In this architecture, state estimation to deal with sensor noise and incomplete or
redundant sensing is a separate process and is as independent as possible
from the control method~\cite{certainty-eq,separation-prin}.
This makes evaluating different control methods much easier.

@@

Closely related control architectures combine Phoneme and Word control,
or combine Phoneme, Word, and Behavior Selection.
This is not a major design change.

@@

Exoskeletons are similar to robots, but there are several important differences.
There are two sets of contact forces: operator-exoskeleton (which we denote
with a subscript $o$ as in $\vf_o$) and exoskeleton-world (which we denote
with a subscript $w$ as in $\vf_w$). If the operator is tightly strapped or
rigidly held by other types of physical constraints, we can lump the operator
body parts with the exoskelton parts which they are attached to in terms
of link inertias, locations of center of mass, and moments of inertia,
and just think of the whole system as one robot.
We can also separate out the operator dynamics if this is more convenient
(we separate them out and ignore them in what follows).
If the operator is loosely connected to the exoskeleton, this introduces a
huge modeling and control problem, and would also be very dangerous for the
operator (as riding a bull or bucking horse in a rodeo is dangerous to a rider).

The dynamics of a revolute exoskeleton are:
\begin{equation}
\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
+ \mJ_w^{\tr}(\vq) \vf_w = \vtau + \mJ_o^{\tr}(\vq) \vf_o
\label{eq:dynamics}
\end{equation}
where $\vq$ are the joint angles and position of the ``root'' of
the exoskeleton, $\dot{\vq}$ and $\ddot{\vq}$ are the corresponding
velocities and accelerations, $\mM$ is the inertia matrix, $\mC()$ are the Coriolis and
centripetal forces, $\mG()$ are the gravitational forces, $\mF()$ are the
friction forces, $\mJ_w^{\tr}(\vq) \vf_w$ (Jacobian matrix multiplied with the
contact forces)
are the exoskeleton-world
contact forces 
expressed as exoskeleton joint torques, 
$\vtau$ are the exoskeleton
joint torques,
and $\mJ_o^{\tr}(\vq) \vf_o$
are the operator-exoskelton forces expressed
as exoskeleton joint torques.
$\vtau$ is augmented with zeros corresponding to the dimensions of $\dot{\vq}$
that correspond to root velocities, since those dimensions are not actuated.

@@

Note that we are ignoring the operator dynamics and operator muscle and tissue
forces resulting in operator joint torques. We will add these in later (or maybe
we won't ???).

We are also ignoring actuator dynamics, which we can handle by augmenting 
the state vector.

[SOMETHING ABOUT MODELING HUMAN + EXOSKELETON + WORLD ]

@@

The goal here is to map from desired motion (in this case desired
accelerations) to exoskeleton actuator commands (and desired operator
and world contact forces).
The exoskeleton actuator torques
attempt to match the torques caused by inertial, Coriolis, centripetal,
gravitational, frictional, and exoskeleton-world contact forces.
\begin{equation}
\vtau = \vtau_{\invdyn} 
= \hat{\mM}(\hat{\vq}) \ddot{\vq}_d
+ \hat{\mC}(\hat{\vq},\hat{\dot{\vq}})
+ \hat{\mG}(\hat{\vq})
+ \hat{\mF}(\hat{\vq},\hat{\dot{\vq}})
+ \hat{\mJ}_w^{\tr}(\hat{\vq}) \hat{\vf}_w
\end{equation}
If there are actuator dynamics or other dynamics, we can include those
dynamics as well.

@@

Note that we cannot change the apparent inertia of the exoskeleton without
some form of acceleration or force feedback.
With acceleration feedback we can make the exoskeleton have a different inertia.
Setting $\ddot{\vq}_d$ to zero and adding acceleration feedback:
\begin{equation}
\vtau = \vtau_{\invdyn} - (\mM_d(\hat{\vq}) - \hat{\mM}(\hat{\vq})) \hat{\ddot{\vq}}
\label{eq:change-inertia1}
\end{equation}
sets the apparent inertia to $\mM_d$ if the dynamic models are perfect, and
the state estimation produces not only perfect positions and velocities but
also perfect acceleration estimates. In the case 
where the operator directly applies torques at the
exoskeleton joints we get:
\begin{equation}
\mM_d (\vq) \ddot{\vq} = \vtau_o
\end{equation}
In robotics this is usually considered a bad idea if $\mM_d$ is smaller (a complex
issue for a matrix) than $\mM$ at any $\vq$. This type of control is vulnerable
to unmodeled dynamics. However, in an exoskeleton, since we have not tried to
cancel the human's dynamics, the human stabilizes a possible unstable exoskeleton
(if the straps are tight enough).

With force feedback we can also make the exoskeleton have a different inertia.
Setting $\ddot{\vq}_d$ to zero and adding force feedback:
\begin{equation}
\vtau = \vtau_{\invdyn} + \mK \hat{\vf}_o
\end{equation}
In the case where the force measurements are perfect and 
the operator directly applies torques at the
exoskeleton joints we get:
\begin{equation}
(\mK + \mI)^{-1} \mM (\vq) \ddot{\vq} = \vtau_o
\end{equation}
By setting $\mK = \mM^{-1}_d (\vq) \mM (\vq) - \mI$ we can achieve
\begin{equation}
\mM_d (\vq) \ddot{\vq} = \vtau_o
\end{equation}

@@

The goal here is to make the exoskeleton imitate a desired linear dynamic system
from the point of view of the operator.
A future white paper will discuss how to make the exoskeleton 
imitate a desired linear dynamic system 
from the point of view of external perturbations.
In the case where there is no force sensing between the operator and the
exoskeleton, the exoskeleton controller maps from exoskeleton configuration
and velocity to desired joint/actuator/synergy forces.
In this case the actuators need to generate forces and torques rather than
positions or angles, or linear or angular velocities.

@@

To make the exoskeleton appear as a locally linear impedance in joint
coordinates to the operator, we add position and velocity feedback to $\vtau$:
\begin{equation}
\vtau = \vtau_{\invdyn} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
\label{eq:impedance}
\end{equation}
so if the dynamic models and state estimation are perfect we get this impedance:
\begin{equation}
\mM \ddot{\vq} - \mK_{\vq} ( \vq - \vq_d ) - \mK_{\dot{\vq}} \dot{\vq} = \vtau_o
\end{equation}
and the operator sees the desirecd impedance.

Although impedance control is extensively utilized in rehabilitation
robotics [18], limited number of studies has been focused on
this method for power augmentation [19, 20].
\begin{verbatim}
[18] R. Riener, L. Lunenburger, S. Jezernik, M. Anderschitz, G. Colombo,
and V. Dietz, "Patient-cooperative strategies for robot-aided treadmill
training: first experimental results," Neural Systems and Rehabilitation
Engineering, IEEE Transactions on, vol. 13, pp. 380-394, 2005.
[19] B.-K. Lee, H.-D. Lee, J.-y. Lee, K. Shin, J.-S. Han, and C.-S. Han,
"Development of dynamic model-based controller for upper limb
exoskeleton robot," in Robotics and Automation (ICRA), 2012 IEEE
International Conference on, 2012, pp. 3173-3178.
[20] W. Yu, J. Rosen, and X. Li, "PID admittance control for an upper limb
exoskeleton," in American Control Conference (ACC), 2011, 2011, pp.
1124-1129.
\end{verbatim}

@@

We can implement operator force control 
if measurements of the operator contact forces with
the exoskeleton are available.
The exoskeleton moves as to create as little force on the operator as possible 
assuming the actuators are torque sources:
\begin{equation}
\vtau = \vtau_{\invdyn} + \mJ_o^{\tr} {\vf_o}_d
+ \mJ_o^{\tr} \mK_{\vf_o} ( \hat{\vf}_o - {\vf_o}_d )
\end{equation}
$\mK_{\vf_o}$ is a gain matrix, $\mJ_o$ is the Jacobian matrix for the operator
force $\vf_o$
Note that this does not require inverting the Jacobian matrix.
We will see that the position control version does invert the Jacobian matrix.

Applying this to force between the operator and the exoskeleton, and assuming
that the operator directly applies exoskeleton joint torques,
\begin{equation}
\vtau = \vtau_{\invdyn} + {\vtau_o}_d - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
\end{equation}
So if our dynamic models and state estimation are perfect, we get:
\begin{equation}
\ddot{\vq} = - \mM^{-1} (\vq) ({\vtau_o}_d + \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d ))
\end{equation}

One can add force damping terms, which sometimes help:
\begin{equation}
\vtau = \vtau_{\invdyn} + {\vtau_o}_d - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
- \mK_{\dot{\vtau}_o} \hat{\dot{\vtau}}_o
\end{equation}

@@

\subsection{Autonomous exoskeleton-world impedance control}

@@

Similarly, we can use exoskeleton-world force control to improve performance,
setting
\begin{equation}
{\vf_w}_d = - \mK_w ( \hat{\vq}_w - {\vq_w}_d ) - \mK_{\dot{w}} \hat{\dot{\vq}}_w
\end{equation}
and add the force error to equation~\ref{eq:impedance}: 
\begin{equation}
\mJ_w^{\tr} \mK_{\vf_w} ( \hat{\vf}_w - {\vf_w}_d )
\end{equation}

The dominant paradigm in force controlled (typically
hydraulic) full size humanoid robots is to handle constraints in inverse dynamics
such as torque and center of pressure limits using online quadratic programming
to solve the inverse dynamics equations. This approach can also be used
to implement and trade off among the many control objectives described so far.

Quadratic programming~\cite{Wikipedia} solves the following problem:
Minimize
\begin{equation}
\vx^{\tr} \mQ \vx + c^{\tr} \vx
\end{equation}
with inequality constraints
\begin{equation}
\mA_1 \vx \le \vb_1
\end{equation}
and equality constraints
\begin{equation}
\mA_2 \vx = \vb_2
\end{equation}
The optimization variable $\vx$ includes accelerations $\ddot{\vq}$, joint
torques $\vtau$, and contact forces $\vf$.
For robots the discrepancy between actual and desired accelerations, joint
torques, and contact forces are all minimized:
\begin{equation}
C_r(\vx) = (\vq - \vq_d)^{\tr} \mW_{\vq} (\vq - \vq_d) + \vtau^{\tr} \mW_{\vtau} \vtau
+ \vf^{\tr} \mW_{\vf} \vf
\end{equation}
$\mW_{\vq}$, $\mW_{\vtau}$, and $\mW_{\vf}$ are weight matrices to trade off the
different errors and error directions.
The dynamics (equation~\ref{eq:dynamics}) are enforced as an equality constraint.
Torque limits, friction limits, center of pressure location constraints,
and unidirectional contact constraints are all
enforced as inequality constraints.
Variants of this optimization
were performed every 1-2ms by various teams to control the
Atlas robots in the DARPA Robotics Challenge.

In the case of exoskeletons, 
part of the optimization criterion could be to
minimize contact forces on the operator:
\begin{equation}
C_e(\vx) = C_r(\vx) + \vf^{\tr}_o \mW_{\vf_o} \vf_o
\end{equation}
Operator force feedback and impedance control could be implemented by specifying a
desired operator force ${\vf_o}_d$ and adding a ``hard'' equality constraint:
\begin{equation}
\vf_o = {\vf_o}_d
\end{equation}
or by adding a ``soft'' constraint in the form of a new term in the optimization
criterion:
\begin{equation}
C_e(\vx) = C_r(\vx) + (\vf_o - {\vf_o}_d)^{\tr} \mW_{\vf_o} (\vf_o - {\vf_o}_d)
\end{equation}

To implement operator impedance control we set the desired operator contact force to:
\begin{equation}
{\vf_o}_d = \mJ_o^{-1} ( \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}} )
\end{equation}
Assuming that the operator directly applies exoskeleton joint torques, this becomes
the equality constraint
\begin{equation}
\vtau_o = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
\end{equation}
or is added to the optimization criterion:
\begin{equation}
C_e(\vx) = C_r(\vx) 
+ (\vtau_o - {\vtau_o}_d)^{\tr} \mW_{\vtau_o} (\vtau_o - {\vtau_o}_d)
\end{equation}
with
\begin{equation}
{\vtau_o}_d = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
\end{equation}

\section{Underactuated and overactuated control}

There are two kinds of underactuation we want to consider. The first
is ``floating body dynamics'', where the robot is not bolted to the floor.
Humanoid robots have floating body dynamics. A ``root'' location (usually somewhere
on the pelvis) is chosen, and the position, velocity, orientation, and angular
velocity of that point and the rigid body it is on
are included in the state vector $\vq$. The corresponding dimensions of the
velocity vector $\dot{\vq}$ are not actuated. We have been sweeping this
issue under the rug so far. It is now useful to be more explicit. A selection matrix
$\mS$ is used to map from actuation variables $\vu$ to joint torques $\vtau$:
\begin{equation}
\vtau = \mS \vu
\end{equation}
The rows in $\mS$ corresponding to the root linear and angular velocities
are all zero.

>>>>

% SIGN ISSUE ON FORCE FEEDBACK o and w: sort it out

@@

\newcommand{\gc}{{\mbox {GravityCompensation}}}
\newcommand{\invdyn}{{\mbox {\tiny ID}}}
\newcommand{\invdynw}{{\mbox {\tiny ID+Fw}}}

@@

This paper surveys possible exoskeleton control architectures and
algorithms.
One goal is to address the question:
{\it How do exoskeleton control approaches compare with 
approaches to humanoid robot control?}
A companion paper surveys implemented exoskeleton control architectures
and algorithms [survey.pdf].

@@

A later white paper will discuss how to handle malfunctions and damage to the
system. Basically this boils down to fault detection, and switchover to
an appropriate safe mode of operation.

A later white paper will address robustness to modeling error and unmodeled
dynamics using simulation.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{design-goals.tex}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Low Level Control}

Low level control presents an API (Application Programmer Interface)
for higher control levels
to access the actuators, sensors, and other robot hardware.
It implements high servo rate processing including feedback control and online
optimization.
This level provides an idealization of the exoskeleton in two ways: 
What degrees of freedom (DOF) are controlled, and how is a degree of
freedom controlled?

\subsubsection{What degress of freedom are controlled?}

The low level control API may directly map on to the exoskeleton hardware
(physical joints,
raw actuators, and raw sensors), or
it may present an ``abstracted'' view or ``virtual'' hardware.
For example, it is often useful to create virtual degrees of freedom that are
actually combinations of joints, or to control points on the robot in Cartesian
or other coordinates. We refer to these virtual degress of freedom as {\bf synergies.}
A complex synergy is the center of mass. Higher level control is made easier by
providing a virtual sensor that measures center of mass location and velocity,
as well as providing a virtual actuator that acts at the center of mass in
a Cartesian fashion (push horizontally and vertically in world coordinates).
On a smaller scale, it may simplify higher level control 
to create virtual actuators that cross
multiple joints. These multi-joint actuators are actually implemented by commands
to individual physical single-joint actuators.

\subsubsection{How is a degree of freedom controlled?}

It is useful for human programmers
to provide them with idealized abstractions of control of a single degree of
freedom. Here are some common alternatives:
\begin{itemize}
\item
Position control. The joint moves to a specified position and stays there independent
of load.
\item
Velocity control. The joint moves at a specified velocity independent of load.
\item
Force or torque control. The joint provides a specified force or torque independent
of its motion.
\item
Impedance control: The user specifies a desired position (for a spring), and commands
a desired
stiffness, damping, and potentially a modified inertia or moment of inertia.
Modifying inertia is more dangerous than providing variable active stiffness
or damping.
Or (almost equivalently),
the user specifies a desired force or torque
which changes according to the position, velocity,
and acceleration of the joint.
\item
Online optimization of accelerations, actuator forces and torques, and contact forces
using quadratic programming. See Section \ref{sec:qp} for further explanation.
\end{itemize}

\subsubsection{Feedback control}

PID or more complex feedback control laws are implemented at this level.

\subsubsection{Feedforward control}

Gravity, friction, actuator dynamics, and other feedforward disturbance
compensation are implemented at this level, as well as any inverse dynamics
or other types of feedforward reference tracking control.

\subsubsection{State estimation}

State estimation is done at this level to compensate for sensor noise.
Disturbance observers for model errors and contact/wind forces
are implemented at this level.

\subsubsection{Existing examples}

Typically low level control is implemented
by a high sampling rate global servo~\cite{Atlas-robot}
or distributed limb or joint-level high sampling rate servos~\cite{Sarcos-robot}.
Often, this controller is provided by the manufacturer of the robot
or exoskeleton and runs on a separate computer system from the
``user's'' or task/application controller~\cite{Sarcos,Atlas}.

\subsection{``Phoneme'' Control}

This level often provides the behavioral primitives or ``verbs'' for robot programming:
providing mid-level controllers
for relatively constant phases of behavior.
These behaviors often involves position or velocity targets (goto or
stay-at X), and trajectory references.
Example phoneme behaviors from walking: left-stance, double-support-1, right-stance, double-support-2, ....
Example behaviors from running: left-foot-down, flight-1, right-foot-down, flight-2, ....
Example behaviors from horizontal jumping: hjump-pushoff, hjump-flight, hjump-prepare-landing, hjump-impact, hjump-balance, ....
Finer or multiple behavioral resolutions may be needed.
For example a walking stance period may
be broken up into heel-strike, first-half, second-half, pushoff, and toe-off.
The first and second halves of stance are divided by when the center of mass
passes over the ankle.
Local inverse kinematics (finding small joint motions for small target changes)
is performed at this level.

\subsection{``Word'' Control}

This level determines the control flow of a behavior: what happens next?
Mid-level controllers are provided for sequences of ``phoneme'' behavioral
units.
This level often involves finite state machines, timers, if-then condition testing,
and gain or behavior
scheduling.
Example word behaviors: walk-to-location(x=7.8m, y=2.3m), run-to-location(speed=3.7m/s),
and horizontal-jump(height=2m, length=5m).
Global inverse kinematics (selection of solution branch) is performed at this level.

\subsection{Behavior Selection}

The highest level of control in the recommended architecture is behavior selection,
which answers the question:
``What should I do next?''
This level selects behaviors and behavioral parameters (targets,
speeds, durations, etc.).
This level often also 
includes decision tables, finite state machines. timers, if-then condition testing, ...
This level often involves a great deal of human operator online interaction,
or is completely done by the operator.

\subsection{State Estimation}

In this architecture, state estimation to deal with sensor noise and incomplete or
redundant sensing is a separate process and is as independent as possible
from the control method~\cite{certainty-eq,separation-prin}.
This makes evaluating different control methods much easier.

\subsection{Error Handling}

Errors must be detected and handled by parameter (target)
adjustment and/or behavior switching
at all of these levels.

\subsection{Why do we recommend this architecture?}

@@

\subsection{What are the alternatives?}

Closely related control architectures combine Phoneme and Word control,
or combine Phoneme, Word, and Behavior Selection.
This is not a major change.

@@

Exoskeletons are similar to robots, but there are several important differences.
There are two sets of contact forces: operator-exoskeleton (which we denote
with a subscript $o$ as in $\vf_o$) and exoskeleton-world (which we denote
with a subscript $w$ as in $\vf_w$). If the operator is tightly strapped or
rigidly held by other types of physical constraints, we can lump the operator
body parts with the exoskelton parts which they are attached to in terms
of link inertias, locations of center of mass, and moments of inertia,
and just think of the whole system as one robot.
We can also separate out the operator dynamics if this is more convenient
(we separate them out and ignore them in what follows).
If the operator is loosely connected to the exoskeleton, this introduces a
huge modeling and control problem, and would also be very dangerous for the
operator (as riding a bull or bucking horse in a rodeo is dangerous to a rider).
We do not address the loosely connected case in this paper.

The dynamics of a revolute exoskeleton are:
\begin{equation}
\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
+ \mJ_w^{\tr}(\vq) \vf_w = \vtau + \mJ_o^{\tr}(\vq) \vf_o
\label{eq:dynamics}
\end{equation}
where $\vq$ are the joint angles of
the exoskeleton, $\dot{\vq}$ and $\ddot{\vq}$ are the corresponding
velocities and accelerations, $\mM$ is the inertia matrix, $\mC()$ are the Coriolis and
centripetal forces, $\mG()$ are the gravitational forces, $\mF()$ are the
friction forces, $\mJ_w^{\tr}(\vq) \vf_w$ (Jacobian matrix multiplied with the
contact forces)
are the exoskeleton-world
contact forces 
expressed as exoskeleton joint torques, 
$\vtau$ are the exoskeleton
joint torques,
and $\mJ_o^{\tr}(\vq) \vf_o$
are the operator-exoskelton forces expressed
as exoskeleton joint torques.

We do not describe how to derive this equation or identify appropriate parameters
for an exoskeleton. We can provide this information if requested.
We are assuming that the equations have been derived taking into account constraints
such as one or two feet on the ground. Until we discuss ``floating body'' dynamics
in Section~\ref{sec:float}, we are assuming contacts do not slip and are not broken.
We are assuming all degrees of freedom (joints) are actuated until 
Section~\ref{sec:float}.

@@

[SOMETHING ABOUT MODELING HUMAN + EXOSKELETON + WORLD ]
Note that we are ignoring the operator dynamics and operator muscle and tissue
forces resulting in operator joint torques. We will add these in later (or maybe
we won't ???).

We are also ignoring actuator dynamics, which we can handle by augmenting 
the state vector and adding additional equations.

@@

The goal here is to map from desired motion (in this case desired
accelerations) to exoskeleton actuator commands (and desired operator
and world contact forces).
The exoskeleton actuator torques
attempt to match the torques caused by inertial, Coriolis, centripetal,
gravitational, and frictional forces.
\begin{equation}
\vtau = \vtau_{\invdyn} 
= \hat{\mM}(\hat{\vq}) \ddot{\vq}_d
+ \hat{\mC}(\hat{\vq},\hat{\dot{\vq}})
+ \hat{\mG}(\hat{\vq})
+ \hat{\mF}(\hat{\vq},\hat{\dot{\vq}})
\end{equation}
If in addition we want to cancel exoskeleton-world contact forces,
we add another term:
\begin{equation}
\vtau = \vtau_{\invdynw} = \vtau_{\invdyn} 
+ \hat{\mJ}_w^{\tr}(\hat{\vq}) \hat{\vf}_w
\end{equation}
If there are actuator dynamics or other dynamics, we can include those
dynamics as well.

@@

Note that we cannot change the apparent inertia of the exoskeleton without
some form of acceleration or force feedback.
With acceleration feedback we can make the exoskeleton have a different inertia.
Setting $\ddot{\vq}_d$ to zero and adding acceleration feedback:
\begin{equation}
\vtau = \vtau_{\invdynw} - (\mM_d(\hat{\vq}) - \hat{\mM}(\hat{\vq})) \hat{\ddot{\vq}}
\label{eq:change-inertia1}
\end{equation}
sets the apparent inertia to $\mM_d$ if the dynamic models are perfect, and
the state estimation produces not only perfect positions and velocities but
also perfect acceleration estimates. In the case 
where the operator directly applies torques at the
exoskeleton joints we get:
\begin{equation}
\mM_d (\vq) \ddot{\vq} = \vtau_o
\end{equation}
In robotics this is usually considered a bad idea if $\mM_d$ is smaller (a complex
issue for a matrix) than $\mM$ at any $\vq$. This type of control is vulnerable
to unmodeled dynamics. However, in an exoskeleton, since we have not tried to
cancel the human's dynamics, the human stabilizes a possible unstable exoskeleton
(if the straps are tight enough).

With force feedback we can also make the exoskeleton have a different inertia.
Setting $\ddot{\vq}_d$ to zero and adding force feedback:
\begin{equation}
\vtau = \vtau_{\invdynw} + \mK \hat{\vf}_o
\end{equation}
In the case where the force measurements are perfect and 
the operator directly applies torques at the
exoskeleton joints we get:
\begin{equation}
(\mK + \mI)^{-1} \mM (\vq) \ddot{\vq} = \vtau_o
\end{equation}
By setting $\mK = \mM^{-1}_d (\vq) \mM (\vq) - \mI$ we can achieve
\begin{equation}
\mM_d (\vq) \ddot{\vq} = \vtau_o
\end{equation}

@@

The goal here is to make the exoskeleton imitate a desired linear dynamic system
from the point of view of the operator.
Section~\ref{sec:impw} discusses how to make the exoskeleton 
imitate a desired linear dynamic system 
from the point of view of external perturbations.
In the case where there is no force sensing between the operator and the
exoskeleton, the exoskeleton controller maps from exoskeleton configuration
and velocity to desired joint/actuator/synergy forces.
In this case the actuators need to generate forces and torques rather than
positions or angles, or linear or angular velocities.

@@

To make the exoskeleton appear as a locally linear impedance in joint
coordinates to the operator, we add position and velocity feedback to $\vtau$:
\begin{equation}
\vtau = \vtau_{\invdynw} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
\label{eq:impedance}
\end{equation}
so if the dynamic models and state estimation are perfect we get this impedance:
\begin{equation}
\mM \ddot{\vq} - \mK_{\dot{\vq}} \dot{\vq} - \mK_{\vq} ( \vq - \vq_d ) = \vtau_o
\end{equation}
and the operator sees the desired impedance.

{\it Although impedance control is extensively utilized in rehabilitation
robotics [18], only a limited number of studies has been focused on
this method for power augmentation [19, 20].}
\begin{verbatim}
[18] R. Riener, L. Lunenburger, S. Jezernik, M. Anderschitz, G. Colombo,
and V. Dietz, "Patient-cooperative strategies for robot-aided treadmill
training: first experimental results," Neural Systems and Rehabilitation
Engineering, IEEE Transactions on, vol. 13, pp. 380-394, 2005.
[19] B.-K. Lee, H.-D. Lee, J.-y. Lee, K. Shin, J.-S. Han, and C.-S. Han,
"Development of dynamic model-based controller for upper limb
exoskeleton robot," in Robotics and Automation (ICRA), 2012 IEEE
International Conference on, 2012, pp. 3173-3178.
[20] W. Yu, J. Rosen, and X. Li, "PID admittance control for an upper limb
exoskeleton," in American Control Conference (ACC), 2011, 2011, pp.
1124-1129.
\end{verbatim}

@@

We can implement operator force control 
if measurements of the operator contact forces with
the exoskeleton are available.
The exoskeleton moves as to create as little force on the operator as possible 
assuming the actuators are torque sources:
\begin{equation}
\vtau = \vtau_{\invdynw} + \mJ_o^{\tr} {\vf_o}_d
+ \mJ_o^{\tr} \mK_{\vf_o} ( \hat{\vf}_o - {\vf_o}_d )
\end{equation}
$\mK_{\vf_o}$ is a gain matrix, $\mJ_o$ is the Jacobian matrix for the operator
force $\vf_o$
Note that this does not require inverting the Jacobian matrix.
We will see that the position control version does invert the Jacobian matrix.

Applying this to force between the operator and the exoskeleton, and assuming
that the operator directly applies exoskeleton joint torques,
\begin{equation}
\vtau = \vtau_{\invdynw} + {\vtau_o}_d - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
\end{equation}
So if our dynamic models and state estimation are perfect, we get:
\begin{equation}
\ddot{\vq} = - \mM^{-1} (\vq) ({\vtau_o}_d + \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d ))
\end{equation}

One can add force damping terms, which sometimes help:
\begin{equation}
\vtau = \vtau_{\invdynw} + {\vtau_o}_d - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
- \mK_{\dot{\vtau}_o} \hat{\dot{\vtau}}_o
\end{equation}

@@

\subsection{Autonomous exoskeleton-world impedance control}
\label{sec:impw}

@@

Similarly, we can use exoskeleton-world force control to improve performance,
setting
\begin{equation}
{\vf_w}_d = - \mK_w ( \hat{\vq}_w - {\vq_w}_d ) - \mK_{\dot{w}} \hat{\dot{\vq}}_w
\end{equation}
and add the force error: 
\begin{equation}
\vtau = \vtau_{\invdyn} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
+ \mJ_w^{\tr} \mK_{\vf_w} ( \hat{\vf}_w - {\vf_w}_d )
\end{equation}

\section{Underactuated and overactuated control and floating body dynamics}
\label{sec:float}

This section describes how to handle underactuation and overactuation using
``floating body dynamics.''
We have been sweeping this
issue under the rug so far. It is now useful to be more explicit.
There are two kinds of underactuation we want to consider. The first
is ``floating body dynamics'', where the robot is not bolted to the floor.
Humanoid robots (and humans) have floating body dynamics.
A ``root'' location (usually somewhere
on the pelvis) is chosen, and the position, velocity, orientation, and angular
velocity of that point and the rigid body it is on
are included in the state vector $\vq$. The corresponding dimensions of the
velocity vector $\dot{\vq}$ are not actuated. A selection matrix
$\mS$ is used to map from actuation variables $\vu$ to joint torques $\vtau$:
\begin{equation}
\vtau = \mS \vu
\end{equation}
The rows in $\mS$ corresponding to the root linear and angular velocities
are all zero.

The second kind of underactuation appears in exoskeletons where we rely on the
operator to actuate a particular joint. For example, many lower body
exoskeltons only actuate the joints in the sagittal plane, leaving lateral
and twist (yaw) joints unactuated or passively actuated.

Passive actuation can be handled by augmenting the
dynamics equations to include the force equation of the passive element,
such as a nonlinear spring.

The underactuation can be handled by reducing the size of the $\vu$ vector
and changing the matrix $\mS$ to reflect the missing actuation.
We can have redundant actuation (more than one actuator across any particular
joint). Here we augment $\vu$ and $\mS$ to accomodate extra actuators.
We can have multijoint actuation, which can be handled in $\vu$ and $\mS$ as well.
